MSE and dynamic state variable model with sigma function

### 2018-04-24 doc/MSE_MAR_18.tex
continued editing text, methods section

The European fisheries management reform of 2013 included the introduction of multi-annual management plans aimed at achieving Maximum Sustainable Yield (MSY) for commericially important fish stocks and the gradual introduction of a landings obligation (LO) \cite{CFP2013}. The LO  should provide incentives for the use of more selective gears and for fisheries to  move away from areas with high levels of unwanted catch \cite{Alzorriz2016, Alzorriz2018}. Achieving single species MSY while landing all commercial catches in fisheries targeting multiple species (mixed fisheries) is challenging because achieving the objective for one species may mean missing the objective for another \cite{Ulrich2017, Batsleer2013}. Forecasting whether such management will be effective depends on understanding the response of individual effort allocation to the management. There is a substantial literature addressing the question of effort allocation in fisheries and the more general question of state-dependent foraging decisions in ecological systems \cite{ClarkandMangel2000,Houston1999}. New tools for state-dependent behaviour of individual fishing vessels, translated into behaviour of the fleet and implemented using stochastic dynamic programming \cite{Alzorriz2018,Batsleer2015, Dowling2011, Poos2010} have been developed. These models generally predict the effect in the short term (within a year) by optimizing an objective function and determine which area best suits this behaviour given the set of incentives that exist (which may also depend on factors such as size, home port, distance to fishing grounds and expected catch rates). Effects of e.g. relocation costs of marine protected areas, the implementation of the landings obligation, have been modelled under these approaches. Using spatial and temporal distributions of catch rates within a single year, these models forecast  high costs in mixed fisheries in the short term as a result of the LO. These costs result from removing the option to adapt landings to quota by discarding parts of the catch \cite{ Alzorriz2018, Batsleer2013,Batsleer2015}. This will result in fishig effort reallocation and early closures of fisheries once quota have been reached. However, the  long term benefits to fisheries that could result from improved selectivity are ignored in those studies

To enable longer term prediction of the effects of changes in fisheires management, one needs to couple the fleet dynamic models that forecast the fleet response to management to  biological dynamic models that forecast the fish population responses to the changing fleet response. Such a link allow to explicilty model the feedback that exists between the fleet dynamics and the populaiton dynamics. Whatever the management regime, the performance of management strategies is conditioned by population dynamics, but also by exploitation dynamics, and particularly by the response of the fleets to management measures. Management strategy evaluation (MSE; \cite{Sainsbury2000, Smith1994, Venables2009}) seeks to study the likely implications of potential harvest policies, or strategies, by evaluating management scenarios on the operation of a mixed fishery and on the fished stocks by including stakeholders and by assessing its robustness to uncertainty \cite{Andersen2010, Kell2007, Prellezo2016, Punt2016}.

MSE involves building a simulation model for the entire process under study, complex enough both to address the current knowledge and uncertainty on the dynamics of fish stocks under fishing pressure, the effect on fishers of variations in stock status and availability, and their responses to those changes and those in management regimes \cite{Venables2009}. At the same time being simple enough to build and calibrate reliably with the available data. More comprehensive MSEs may also take the effects of the fishery on the broader environment, and economic and social effects into account \cite{Dichmont2008, Fulton2007}. In such modelling framework, advice and management systems can be quantitatively evaluated through simulation testing \cite{Dichmont2013, Kell2007}.

We present an MSE that encompasses the individual harvester decision-making process and socioeconomic drivers on management effectiveness, i.e. if the fleet dynamic model results in a short term movement of vessels to allocate effort both spatially and temporally, we also expect longer term changes to occur in fish populations and the economy of the fishery \cite{Alzorriz2018}. Such combination needs to be responsive to all the drivers that influence and motivate the actual fleet and to respect existing constraints \cite{Venables2009}. The drivers will include the (real or perceived) local abundance and catchability of the fished stocks and various resource costs, principally that of fuel. The constraints will include the obvious management-imposed constraints such as levels of total allowable catch, and economical constraints such as the contribution of the annual fines for exceeding catch quotas.

@@

A Managament Strategy Evaluation framework was used to forecast the dynamics of the mixed fishery. The essential elements of the framework are summarized in Figure \ref{fig:MSE}. The framework consists of an Operating Model (OM) and a Management Procedure (MP). The OM captured key processes in the dynamics of the fish populations, and can be thought of as a minimum realistic model \cite{Punt1995}. This resulted in simulated measurements (called "observation model") such as individual harvester decision-making (including error) and the consequent biomass of fish stocks, including the essential elements for calculating the individual harvester economic performance. The observation data were then passed to the "management model" (section \ref{sec2.3}). The management model encompassed the harvest control rules (HCR). The HCR referenced to biological reference points (Fmsy) to produce management actions in the form of a harvest or effort level: changes in selectivity or spatial and temporal reallocations or restrictions of fishing effort. Costs and benefits, both economic and in terms of risks to both stock and livelihoods, could be computed and compared across different scenarios and management procedures (section \ref{sec2.3}).

@@

For purposes of model simplicity, we considered a hypothetical mixed fishery on two species. The fishery was modelled in a individual based dynamic state variable model. In order to allowed for fishing effort allocation, the model was seasonally (6 seasons) and spatially (areas A and B) explicit. The dynamics of the two fish species were governed by a simple age-structured model that is also seasonally and spatially explicit. The age structured model for the fish species was based on 4 cohorts, and individuals were assumed to be born at age 0, and in season 1. The number of fish $i$ of age $a$ at year $y$, in season $s$ and area $p$ was written as $N_i (a, y, s, p)$. The number of new-born individuals in the model was independent of the size of the adult population, but was function of area $p$ ($R_i (p)$). Numbers of recruits were set to 100 individuals during the first season, and spawning population of species 1 ocurred in area $A$, while species 2 did so in area B. Thus, individuals could be born in each of the areas, such that:

@@

Moreover, combining the weight-based von Bertalanffy growth equation ($L\infty= 20$, $k= 0.093$), with the age-weight relationship ($W= L\infty*(1-e^{-k*a})$); weight per fish was higher for the older fish. Mortality in the model could occur from catches in the fishery, and natural causes such as predation, diseases, and scenesens. However, for simplicity we assumed that his natural mortality was negligible. Althought such absence of natural mortality impossible in reality, one could argue that the model thus mimics long-lived species. The decrease in population numbers was thus the result of the catches ($C_i (a, y, s, p)$), which were in turn a function of age dependent catchability $q(a)$, and fishing effort $E (y,s,p)$ in any time, season, and area.

\begin{equation}
C_i (a, y, s, p) = q(a) * E(y,s,p),
\end{equation}

and

\begin{equation}
N_i (a, y, s, p) = N_i (a, y, s, p) - C (a, y, s, p). 
\end{equation}

It was assumed that fishing mortality ocurred before migration ocurred. Migration for each species was defined by a matrix $M$ that defines the impact of migration (.e. immigration-emigration) on a given stock was also dependent on the relative sizes of the stocks. The size of that matrix  was defined by the number of age classes, seasons and number of immigrants into area $p$ ($im$) and emigrants leaving area $p$ ($em$) in the model. A 20 \% constant migration during any season was set between areas. The migration in population numbers ($M (a, 1, s, im, em)$) was thus the proportion ($P (a, y, s, i, em, im)$) of the population available or vulnerable to fishing after migration occurs:

\begin{equation}
M (a, 1, s, im, em) = N (a, y, s, em) * P (a, y, s, i, em, im),
\end{equation}

@@

@@

>>>>

Management strategy evaluation (MSE; \cite{Sainsbury2000, Smith1994, Bunnefeld2011}) seeks to study the implications of management strategies using simulation\cite{Punt2016}. Such simulation should include all important processes of a fishery, which is inherently a socio-ecological system, with coupled dynamics of the fishing fleets, the exploited stocks, and their governance\cite{Punt2016,Rasemeyer2007}. Given the complexity of such a socio-ecological system the uncertainy about the dynamics and feedbacks are consideable. As a result, any forecasted success in achieving goals in a management strategy may thus be inperceivably small compared to the stochasticity in results generated in the system, or worse, be counteracted by unintended consequences of the management strategy.The robustness of management strategies to the uncertainy in the processes that govern fisheries systems thus need to be accounted for \cite{Andersen2010, Kell2007, Prellezo2016, Punt2016}.

The adaptive response of fishers to their environment is one of the key uncertainties requiring attention in MSE \cite{Fulton2007}. Simulation tools for such adaptive responses are available in fisheries, borrowing methodology from the more general question of state-dependent foraging decisions in ecological systems \cite{ClarkandMangel2000,Houston1999}. New tools for state-dependent behaviour of individual fishing vessels, translated into behaviour of the fleet and implemented using stochastic dynamic programming \cite{Alzorriz2018,Batsleer2015, Dowling2011, Poos2010, Gillis1995} have been developed. These models generally predict the effect in the short term (within a fishing trip, or a quota year) by optimizing a utility function and determine which choices best yields the best chance of increasing utility, while keeping track of the state of each individual. The effect of a choice on the utility depends on the economic environment,   such as the home port of the vessel and the distance to fishing grounds, and the bioological enviroment, such as the spatial distribution of the resources. Using such dynamic variable state models, effects of e.g. relocation costs of marine protected areas have been modelled \cite{Dowling2011}.

Dynamic state variables have also been used to describe discarding behaviour \cite{Gillis1995, Batsleer2013, Batsleer2015}. These studies may shed light on the potential outcomes of the European fisheries management reform of 2013. That reform included using the use of the Maximum Sustainable Yield (MSY) reference points as targets for exploiting commericially important fish stocks and the gradual introduction of a landings obligation (LO) \cite{CFP2013}. Indeed, using spatial and temporal distributions of catch rates within a single year, dynamic state variable models forecasted  high costs in mixed fisheries in the short term as a result of the LO. These costs result from removing the option to adapt landings to quota by discarding parts of the catch \cite{ Alzorriz2018}. This will result in fishig effort reallocation and early closures of fisheries once quota have been reached. However, the  long term benefits to fisheries that could result from improved selectivity are ignored in those studies. Ignoring the potential for improved selectivity leaves out a key element in the perceived benefit of the LO. Because the LO should provide incentives for the use of more selective gears and for fisheries to  move away from areas with high levels of unwanted catch \cite{Alzorriz2016, Alzorriz2018}, the improved selectivity should lead to higher long term catches. Achieving single species MSY while landing all commercial catches in fisheries targeting multiple species (mixed fisheries) is challenging because achieving the objective for one species may mean missing the objective for another \cite{Ulrich2017, Batsleer2013}. Forecasting whether such management will be effective depends on understanding the response of individual effort allocation to the management.

To enable longer term prediction of the effects of changes in fisheries management, and thus complete the Management Strategy Evaluation, one needs to couple the fleet dynamic models that forecast the fleet response to management to  biological dynamic models that forecast the fish population responses to the changing fleet response. We present an MSE that encompasses the individual harvester decision-making process and socioeconomic drivers on management effectiveness, i.e. if the fleet dynamic model results in a short term movement of vessels to allocate effort both spatially and temporally, we also expect longer term changes to occur in fish populations and the economy of the fishery \cite{Alzorriz2018}. Such combination needs to be responsive to all the drivers that influence and motivate the actual fleet and to respect existing constraints \cite{Venables2009}. The drivers will include the (real or perceived) local abundance and catchability of the fished stocks and various resource costs, principally that of fuel. The constraints will include the obvious management-imposed constraints such as levels of total allowable catch, and economical constraints such as the contribution of the annual fines for exceeding catch quotas.

@@

The Managament Strategy Evaluation framework was used to forecast the dynamics of a hypothetical mixed fishery on two species. The framework consists of an Operating Model (OM) and a Management Procedure (MP). In the fishery, there are three essential elements: (i) a collection of size strucured fish stocks, whose dynamics are governed by annual reproduction, growth, migration, and mortality. (ii) A management body that evaluates the fishing pressure and aims to set annual quotas in accordance with fishing mortalities dictated by FMSY reference points. (iii) a fleet of individual fishers who aim to make the best use of their annual quota. Within this mixed fishery, individual vessels make adaptive choices on fishing location and discarding that depend on the distribution of the resources, and the quota that they are allocated. Each of these elements is discussed in more detail below. The essential elements of the framework are summarized in Figure \ref{fig:MSE}.

The OM captured the key processes in the dynamics of the fish populations, the fisheries and the management body, and can be thought of as a minimum realistic model \cite{Punt1995}.  The operating model thus includes  individual harvester decision-making (including error) and the consequent biomass of fish stocks, including the essential elements for calculating the individual harvester economic performance. The  management body used the MP to make its decision on how to respond to the state of the resource. This management procedure thus includes the data collection from the fishery, how these are interpreted, and the Harvest Control Rule that dictates the limits that are set to the fishery.

@@

In order to allow the fishery to make spatial and temporal choices, the model was seasonally (six seasons) and spatially (two areas) explicit. The dynamics of two fish stocks were modelled by an age-structured model that is also seasonally and spatially explicit. The maximum age in the age-structured fish model was 4 years. The number of fish of stock $i$ of age $a$ at year $y$, in season $s$, and area $p$ was written as $N_i (a, y, s, p)$. The two stocks are named "stock 1" and "stock 2". and the areas are named "area a" and "area b". Individuals were born at age 0, at the start of each year $y$, in season 1. The number of new-born individuals in the model was independent of the size of the adult population, but was function of area $p$, and written as $R_i (p)$. Hence,

@@

In the example used for our simulations, it was assumed that there were 100 new-borns per year for both stocks. The difference between the stocks is that all new-borns of stock 1 were born in "area a", while all new-borns of stock 2 were born in "area b".

Mortality in the model resulted only from catches in the fishery. Mortality from natural causes such as predation, diseases, and scenesens was assumed to be negligible in comparison. Althought such absence of natural mortality impossible in reality, one could argue that the model thus mimiced long-lived species. The decrease in population numbers was thus the result of the catches ($C_i (a, y, s, p)$), which were in turn a function of age dependent catchability $q_i (a)$, and fishing effort $E (y,s,p)$ in any time, season, and area. Thus,

\begin{equation}
C_i (a, y, s, p) = N_i (a, y, s, p) * q(a) * E(y,s,p).
\end{equation}

The mortality reduces the population numbers as for seasons 2 to 6 as

\begin{equation}
N_i (a, y, s+1, p) = N_i (a, y, s, p) - C_i (a, y, s, p), 
\end{equation}

while in season 1, the population numbers were defined as

\begin{equation}
N_i (a+1, y, 1, p) = N_i (a, y, 6, p) - C_i (a, y, 6, p). 
\end{equation}

Migration for each species was defined by a matrix $M_i$ that defines the impact of migration (.e. immigration-emigration) on a given stock was also dependent on the relative sizes of the stocks. The size of that matrix  was defined by the number of age classes, seasons and number of immigrants into area $p$ ($im$) and emigrants leaving area $p$ ($em$) in the model. A 20 \% constant migration during any season was set between areas. The migration in population numbers ($M_i (a, 1, s, im, em)$) was thus the proportion ($P (a, y, s, i, em, im)$) of the population available or vulnerable to fishing after migration occurs:

\begin{equation}
M_i (a, 1, s, im, em) = N_i (a, y, s, em) * P (a, y, s, i, em, im),
\end{equation}

@@

Moreover, combining the weight-based von Bertalanffy growth equation ($L\infty= 20$, $k= 0.093$), with the age-weight relationship ($W= L\infty*(1-e^{-k*a})$); weight per fish was higher for the older fish.

@@

The management model encompassed the harvest control rules (HCR). The HCR referenced to biological reference points (Fmsy) to produce management actions in the form of a harvest or effort level: changes in selectivity or spatial and temporal reallocations or restrictions of fishing effort. Costs and benefits, both economic and in terms of risks to both stock and livelihoods, could be computed and compared across different scenarios and management procedures (section \ref{sec2.3}).
